Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
updated 2.2

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy within our sample by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\bf enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] +\nabla\cdot(\kappa \nabla T)
, 
\label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature,
pressure and specific entropy of the gas, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p =
\rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$. {\bf BDM: take $\mu = 0.62$ here for fully ionized gas}

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years. $\eta\simeq 0.02 \left(\frac{t}{th}\right)^{-1.3}$ 
(see, e.g., \citealt{Ciotti+91})\footnote{Note that 
\citealt{Ciotti+91} gives the overall mass return rate from evolved
stars as a function of B-band luminosity. We give the mass return rate
per unit volume. However, our expressions are equivalent.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy relative to the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

The second source term in equation (\ref{eq:dsdt}) accounts for
heating transportation by electron conductivity. We take the
conductivity to be $\kappa_{\rm eff}=\kappa_{\rm spitzer}/(1+\psi)$,
as described in \citealt{DaltonBalbus:1993a}. $\kappa_{\rm spitzer}$
is the classical \citealt{Spitzer62} value, $\kappa_{\rm spitzer} =
\kappa_0 T^{5/2}$ ($\kappa_0\simeq 2\times 10^{-6}$ in cgs units). The
term in the denominator is a flux limiter. The conductive flux should
saturate when the mean free path for electron coulomb scattering
becomes comparable to the temperature length scale.  More precisely
$\psi=\frac{\kappa_{\rm spitzer} \nabla T}{5\phi \rho c_s^3}$ (the
Spitzer conductive flux over the saturated conductive flux; $\phi$ is
a dimensionless constant of order unity or less). Although even a weak
magnetic field could in principle strongly suppress the conductivity
perpendicular to the field, the radially-decreasing temperature
profiles of interest are susceptible to the magneto-thermal
instability (\citealt{Balbus01}), the non-linear evolution of which is
believed to result in a radially-directed field structure
(\citealt{Parrish&Stone07}).

Following \citet{ShcherbakovWong+:2014a}, the term in the entropy
equation $\propto \vw^2 = \sigma(r)^2+v_{w,0}^2$ accounts for external
heating sources.  Here $\sigma(r) = \sqrt{GM_{\rm enc}/r}$ is the
velocity dispersion of the stars, thus accounting for the fact that at
a minimum the gas experiences shock heating between the stellar winds
of velocities moving with respect to each other a characteristic
velocity $\sim \sigma$.  This heating is present if the wind velocities themselves
are low, $\ll \sigma$, as characterize the majority of AGB mass loss.
The term $v_{w,0}^{2}$ parameterizes additional sources of heating,
such as main sequence winds, millisecond pulsars, supernovae, etc.  As
discussed in $\S\ref{sec:heating}$, the value of $v_{w,0}$ is
uncertain and will in general depend on the SMBH mass and the age of
the stellar population.

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability (\citealt{Brandenburg:2003a}).  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

We do not include radiative cooling in the entropy equation because
across much of the parameter space radiative cooling is negligible
compared to other sources of cooling (e.g., advection).  Although
radiative cooling cannot be neglected in some cases (especially if
$\vwO$ is low or $\eta$ is high), our 1D scheme cannot account for the
true multi-phase behavior of local thermal instabilities anyways
(e.g.~\citealt{McCourt+12}).  Our steady state solutions neglecting
radiation can nevertheless be used to ascertain over what parameter
space cooling instabilities are likely to develop
($\S\ref{sec:cooling}$).
%Dependence of the heating to cooling ratio has a far steeper
%dependence on vwO than on eta...

>>>>

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy model by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\rm enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] +\nabla\cdot(\kappa \nabla T)
, 
\label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature,
pressure and specific entropy of the gas, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p =
\rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$. {\bf BDM: take $\mu = 0.62$ here for fully ionized gas - what did you do?}

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t/t_h)^{-1.3}$ at time $t$ following a burst of star formation (e.g., \citet{Ciotti+91}),\footnote{Note that 
\citet{Ciotti+91} gives the overall mass return rate from evolved
stars as a function of B-band luminosity. We give the mass return rate
per unit volume. However, our expressions are equivalent.} although $\eta$ will generally be higher in cases of ongoing star formation.

Source terms $\propto q$ also appear in the momentum and entropy equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) due to the fact that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy relative to the mean flow (see also below).  Physically these terms result from the mismatch between the properties of virialized gas injected locally by stellar winds and the mean properties of the background flow.   As we shall discuss in $\S$, the cooling term $\propto q p/\rho$ represents an important stabilizing term in comparison to radiative cooling.

The last term in equation (\ref{eq:dsdt}) accounts for heat transportation by electron conductivity, where $\kappa_{\rm eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free path for electron coulomb scattering becomes comparable to the temperature length scale.  Here $c_s \equiv (kT/\mu m_p)^{1/2}$ is the isothermal sound speed and $\phi$ is an uncertain dimensionless constant, typically taken to be of order unity of less (we adopt $\phi = XXX$).  Even a weak magnetic field could in principle also suppress the conductivity perpendicular to the field.  However, the radially-decreasing temperature profiles of interest are susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field structure (\citealt{Parrish&Stone07}).

Following \citet{ShcherbakovWong+:2014a}, the term $\propto \vw^2 = \sigma(r)^2+v_{w,0}^2$ in the entropy equation accounts heating sources.  Here $\sigma(r) = \sqrt{GM_{\rm enc}/r}$ is the velocity dispersion of the stars, accounting for the minimum amount of shock heating from the stellar winds due to the random movement of stars with respect to each other at the characteristic velocity $\sim \sigma$.  This heating is present even for low wind velocities $\ll \sigma$, as characterize the majority of AGB mass loss.
The term $v_{w,0}^{2}$ parameterizes the contribution of additional sources of heating, including main sequence winds, millisecond pulsars, supernovae, AGN feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of $v_{w,0}$ is uncertain and will in general depend on the SMBH mass and the age of the stellar population.

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability (\citealt{Brandenburg:2003a}).  We solve the equations for
different choices of $v_{\rm w,0} = 300, 600, 1200$ km s$^{-1}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM structure, the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic inflow on small radial scales and a possible supersonic outflow on large scales.

Radiative cooling is not included in the entropy equation, although its effect is assessed critically in $\S\ref{sec:cooling}$.  Across much of the parameter space radiative cooling is negligible compared to other sources of cooling (e.g., advection).  Radiative cooling cannot be neglected in some cases (if
$\vwO$ is low and/or $\eta$ is high), once it becomes important it leads to thermal instabilities, which our one-dimensional time-dependent modeal cannot account for anyways (e.g.~\citealt{McCourt+12}).  A chief goal of this work is instead to use our steady state solutions neglecting radiation to ascertain over what physical conditions cooling instabilities are likely to develop ($\S\ref{sec:cooling}$).
